---
title: 'Part 1: Interactive RMarkdown'
author: "Allison Horst"
output: html_notebook
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

To create this document, make sure to choose the "Shiny document" option when you create the file! That will add the `runtime: shiny` line to your YAML, and give you some nice examples you can check out. 

Once you create the new RMarkdown Shiny document, click 'Run Document' up top to see what appears. An interactive document where users can change what they see! 

How cool is this for your boss, client, etc? Some of the benefits of an interactive app, without the bulk -- this is a nice option for simple & sleek documents where you want the user to have some flexibility about variables, etc. that appear in a graph or table. 

Let's make something new. 

### Attach packages

In the setup chunk, attach:

- `tidyverse`
- `palmerpenguins`

### Add some interactivity! 

```{r flippers, echo=FALSE}
inputPanel(
  selectInput(inputId = "pick_spp", label = "Pick penguin species:",
              choices = c("Adélie" = "Adelie", "Chinstrap", "Gentoo"), selected = "Adelie"),
  
  sliderInput("pt_size", label = "Set point size:",
              min = 1, max = 10, value = 5, step = 0.5)
)

penguin_set <- reactive({
  penguins %>% 
  filter(species == input$pick_spp)
})

renderPlot({
  ggplot(data = penguin_set(), aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(size = input$pt_size)
})
```

...which is pretty cool. 

You can publish to a free [shinyapps.io](https://www.shinyapps.io/) account! 
