---
title: 'Part 2: tsibble basics for time series exploration'
author: "Allison Horst"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(slider)
```

## 1. Always look at your data

Toolik Station (LTER) meteorological data (Source: Source: Shaver, G. 2019. A multi-year DAILY file for the Toolik Field Station at Toolik Lake, AK starting 1988 to present. ver 4. Environmental Data Initiative.)

Notice that the date parsed (assumed class) as *character.* That limits the nice time series features we can use, so we'll quickly convert it into a tsibble (a time series data frame) so that we can use functions in `feasts` and `fable` to explore & analyze it. 

### Read in data: 

```{r}
toolik <- read_csv(here("data","toolikweather.csv"))
```

### Convert the data frame to a tsibble

Go ahead and try plotting the data as imported. 

```{r, eval = FALSE}
ggplot(data = toolik, aes(x = date, y = mean_airtemp)) +
  geom_line()

# Booo we get a warning (only one observation per series)
```

Notice that it doesn't work - because R doesn't understand the date is a *date* until we tell it.

Let's go ahead and convert it to a tsibble using the `as_tsibble()` function. First, we'll need to convert the date to a `date` class, *then* convert to a tsibble:

```{r}
toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  as_tsibble(key = NULL, index = date)
```

Now let's plot it: 
```{r}
ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line() +
  labs(x = "Date",
       y = "Mean daily air temperature (Celsius)\n at Toolik Station")
```

We need to ask some big picture questions at this point, like: 

- Does there appear to be an overall trend? No.
- Does there appear to be seasonality? Yes.
- Does there appear to be cyclicality? Unsure.
- Any notable outliers or additional patterns? No noted.

## 2. Use `index_by()` to aggregate time series by increments

We will use `index_by()` instead of `group_by()` to do the trick. See `?index_by()` to group by a time index, then `summarize()` to specify what to calulate & return for each interval. 

```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% 
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))
```

Now let's take a look: 
```{r}
ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() 

# Or break it up by month: 
toolik_month %>% 
  ggplot(aes(x = year(yr_mo), y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE)) +
  labs(x = "Year",
       y = "Annual mean air temperature (Celsius)",
       title = "Toolik Station mean annual air temperature",
       subtitle = "1988 - 2018",
       caption = "Source: Shaver, G. 2019. A multi-year DAILY \nweather file for the Toolik Field Station at Toolik Lake, AK\n starting 1988 to present. ver 4. Environmental Data Initiative.")
```

Can you do other increments with `index_by()`? Absolutely! **See `?index_by()` for grouping options!**

Let's find the yearly average for 2000: 

```{r}
toolik_annual <- toolik_ts %>% 
  index_by(yearly = ~year(.)) %>% 
  summarize(annual_airtemp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_annual, aes(x = yearly, y = annual_airtemp)) +
  geom_line()
```

And how about a weekly average?
```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_weekly, aes(x = weekly, y = weekly_airtemp)) +
  geom_line()
```
## 3. Use `filter_index()` to filter by date-times!

We can use `filter_index()` specifically to help us filter data by time spans. See `?filter_index()` for more information.

**Example 1:** Filter from June 2000 through October 2001

```{r}
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")
```

**Example 2:** Filter from April 10, 2006 to May 15, 2006

```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")
```

**Example 3:** Filter from December 20, 2017 to the end of the dataset

```{r}
toolik_ts %>% 
  filter_index("2017-12-20" ~ .)
```

## 4. Explore changes in seasonality with seasonplots

Let's look at seasonality over the years with a seasonplot, using the `feasts::gg_season()` function. Notice that we can still do wrangling on a tsibble like we would with a normal data frame: 

```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% 
  gg_season(y = mean_airtemp)
```

Daily measurements seems a bit excessive to return in this visualization, right? Maybe it makes more sense to use the monthly averages in ``. 

```{r}
# Now a season plot: 
toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  theme_minimal() +
  labs(x = "Year",
       y = "Mean monthly air temperature (Celsius)",
       title = "Toolik Station air temperature")
```

## 5. Seasonal subseries plots

Sometimes it can be useful to explore how values within one season/month/etc. change over time (e.g. across years). 

We can use `gg_subseries()` to explore how values change within a specified window over time. 

Do you notice any trends that differ across the months? 

```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)
```

## 6. Moving averages in tsibbles

We'll use the `slider` package to find moving (or rolling) averages for different window sizes. 

The general structure will tend to be something like: 

`df %>% slide(variable, function, .before = , .after = )`

Let's make a test vector just so we can see how this works: 

```{r}
set.seed(2021)
test<- rnorm(100, mean = 40, sd = 10)

# Show the series based on values +2 and -2 from each observation
# Use ~.x to show the windows
slide(test, ~.x, .before = 2, .after = 2)

# Change that to a function name to actually calculate something for each window
# Note that I add `as.numeric` here, since the outcome is otherwise a list
w5 <- as.numeric(slide(test, mean, .before = 2, .after = 2))
w5

# Find the mean value of a window with n = 11, centered:
w11 <- as.numeric(slide(test, mean, .before = 5, .after = 5))
w11

# Find the mean value of a window with n = 19, centered:
w19 <- as.numeric(slide(test, mean, .before = 9, .after = 9))
w19

# Plot these together: 
combo <- data_frame(time = seq(1:100), test, w5, w11, w19)

ggplot(data = combo) +
  geom_line(aes(x = time, y = test)) +
  geom_line(aes(x = time, y = w5), color = "red") +
  geom_line(aes(x = time, y = w11), color = "blue") +
  geom_line(aes(x = time, y = w19), color = "orange")

```


Now for an example with our Toolik Station data, let's say we want to find the *average* value at each observation, with a window that extends forward and backward n days from the observation:

```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, mean, .before = 7, .after = 7)))

roll_toolik_61 <- toolik_ts %>% 
  mutate(ma_61d = as.numeric(slide(toolik_ts$mean_airtemp, mean, .before = 30, .after = 30)))

ggplot() +
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp), size = 0.2, color = "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d), color = "orange") +
  geom_line(data = roll_toolik_61, aes(x = date, y = ma_61d), color = "red") +
  theme_minimal()

```

## 7. Autocorrelation function

We'll look at outcomes for both daily lags (yikes) and monthly lags (cool).

```{r}
toolik_ts %>%
  ACF(mean_airtemp) %>%
  autoplot()

toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
```

##8. Decomposition

Here we will use STL decomposition (Seasonal, Trend, and Loess) decomposition. You can read about the advantages of STL decomposition here: https://otexts.com/fpp2/stl.html.


```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

components(toolik_dec) %>% autoplot()
```

## END Part 2
